Comparing R packages for calculation of persistent homology

### 2020-03-19 tdabench-paper/tdabench.Rmd
updated intro

Topological data analysis (TDA) is a broad set of methodologies that characterize structural features of data sets inspired by topological principles.
Within the umbrella of TDA, persistent homology represents an algebraic approach to understanding the number, characteristics, and persistence of structural features in an n-dimensional point cloud.

In the basic workflow of persistent homology, a series of simplicial complexes are generated on point clouds to characterize the topological features. 
Simplicial complexes are connected sets of points, lines, and n-dimensional shapes.
There are several methods to generate these complexes on point clouds. 
In this paper, we are using the Vietoris-Rips and alpha complex for our persistent homology calculations. 
In both methods, a series of data structures, called simplicial complexes, are built on top of a point cloud.
These complexes can be viewed as approximating real topologic relationships in the data.
The exact method of constructing these complexes is described in the Mathematics section. 
However, the key takeaway is that the simplicial complex attempts to reveal the underlying topology of the point cloud. 
These features can be described by dimension and "persistence." 
Long lasting (persistent) features that exist across the series of simplicial complexes intuitively represent "real" geometrical patterns as opposed to noise.

One limitation of persistent homology is its run time limits for processing large numbers of points ( $> 10^9$ points) and calculating of high dimensional features ($>$ 2 dimensions). 
Alpha complex calculations have a complexity of $O(n^\frac{d}{2})$ and Vietoris-Rips complex calculations have a complexity of $O(2^n)$ where n is the number of points and d is the dimension of the point cloud \citep{roadmap}. The limits and computational resources needed to run complex data sets with major persistent homology C++ software have been well characterized \cite{roadmap}.

There are several C++ libraries that calculate alpha and Vietoris-Rips complexes.
Dionysus provides alpha and Vietoris-Rips complex functionality \cite{Dionysus}.
GUDHI is another C++ library that also provides both functionalities as well as other TDA methodologies \cite{gudhi}. These libraries have been implemented in R with the TDA package \cite{TDA} providing an easier interface to utilize persistent homology calculations.

A new C++ library, Ripser, was developed as a more efficient way to calculate Vietoris-Rips complexes. The code is supposed to outperform other similar libraries (e.g. Dionysus and GUDHI) by a factor over 40 in runtime \cite{Ripser}. 
This library was given an R interface in the package TDAstats \cite{TDAstats}. 
TDAstats represents a more specialized package for topological data analysis compared to the TDA package.
It calculates persistent homology using the more efficient Ripser algorithm and provides data visualization with integration of ggplot. While Ripser is known to be faster than the other libraries for Vietoris-Rips complex calculations in C++, how this translates in the R interface is not clear.
While C++ generally runs faster than R, R provides higher level functionality, and appeals to a broad audience of data science users.

Topological data analysis has had a broad range of usage from viral evolution to physical chemistry\cite{TDA-Viral,TDA-PChem}. 
Given that persistent homology can apply to any n-dimensional data set, it has wide ranging applicability. 
Its main limiting factor is computational for calculating high dimensional features in large multi-dimensional data sets. 
We believe lack of analysis of these higher dimensional features may under utilize persistent homology's analysis potential.
In this project, we explore different methodologies to find efficient options that could solve this issue. 
We specifically aim to characterize the computational performance (run time) and limits of persistent homology calculations between Vietoris-Rips complexes in the \CRANpkg{TDAstats} and \CRANpkg{TDA} R packages.
We also explore the differences in computational performance between Vietoris-Rips and alpha complexes in R to assess which datasets are most appropriate for each algorithm.

\section{Mathematics of Persistent Homology}
This section summarizes the mathematics behind generating simplicial complexes on point clouds in the workflow of persistent homology. 
An n-dimensional simplex is the convex hull of n+1 vertices.
More intuitively, n-dimensional simplex is the simplest n-dimensional object (e.g. a 1 simplex is a line, a 2-simplex is a triangle, 3-simplex is a tetrahedron, etc.).
These simplices can be glued together on a common subsimplices to form a simplicial complex (e.g. two triangles sharing a common side).
In a simplicial complex, certain features will arise that can be characterized by dimension number (Betti number): 0-dimensional features refer to connected components, 1-dimensional features refer to holes, 2-dimensional features refer to cavities, etc. \cite{phom-survey}

@@

In both methods, the boundary matrix records all simplicial complexes for a given parameter value. 
From this matrix, the persistent homology matrix is calculated. 
The size of the boundary matrix can describe the memory complexity on the RAM for persistent homology calculations. 
We compare this memory complexity between alpha and Vietoris-Rips complexes for this project.

>>>>

Topological data analysis (TDA) is a broad set of methodologies that characterizes structural features of datasets inspired by topological principles.
It has a broad range of usage from viral evolution to physical chemistry \citep{TDA-Viral,TDA-PChem}. 
Within the umbrella of TDA, persistent homology represents an algebraic approach to understanding the number, characteristics, and persistence of structural features in an $n$-dimensional point cloud.
In the basic workflow of persistent homology, a series of simplicial complexes are generated on point clouds to characterize topological features.
There are several methods to generate these complexes on point clouds. 
In this paper, we focus on persistent homology of the Vietoris-Rips and alpha complexes, which use simplicial complexes to approximate topologic relationships in point clouds.
The exact method of constructing these complexes is described in the Mathematics section.
Persistent features more likely represent real geometric patterns, rather than noise.

There are several C++ libraries available to researchers that calculate alpha and Vietoris-Rips complexes, such as Dionysus, GUDHI, and Ripser \citep{Dionysus,gudhi,Ripser}.
These libraries have been wrapped in R by the \CRANpkg{TDA} and \CRANpkg{TDAstats} packages \citep{TDA,TDAstats}.
Although useful, calculating persistent homology for large datasets is often limited due to computational complexity \citep{roadmap}.
As a result, researchers often limit persistent homology analysis to lower dimensions.
However, ignoring features in higher dimensions may cause significant information loss, underutilizing persistent homology's capabilities.
Here, we aim to benchmark two R packages - \CRANpkg{TDA} and \CRANpkg{TDAstats} - and enable researchers to most efficiently calculate persistent homology in R.

\section{Mathematics of Persistent Homology}

This section summarizes the mathematics behind generating simplicial complexes on point clouds in the workflow of persistent homology. 
An n-dimensional simplex is the convex hull of n+1 vertices.
More intuitively, n-dimensional simplex is the simplest n-dimensional object (e.g. a 1 simplex is a line, a 2-simplex is a triangle, 3-simplex is a tetrahedron, etc.).
These simplices can be glued together on a common subsimplices to form a simplicial complex (e.g. two triangles sharing a common side).
In a simplicial complex, certain features will arise that can be characterized by dimension number (Betti number): 0-dimensional features refer to connected components, 1-dimensional features refer to holes, 2-dimensional features refer to cavities, etc. \cite{phom-survey}

@@

In both methods, the boundary matrix records all simplicial complexes for a given parameter value. 
From this matrix, the persistent homology matrix is calculated. 
The size of the boundary matrix can describe the memory complexity on the RAM for persistent homology calculations. 
We compare this memory complexity between alpha and Vietoris-Rips complexes for this project.
Alpha complex calculations have a complexity of $O(n^\frac{d}{2})$ and Vietoris-Rips complex calculations have a complexity of $O(2^n)$ where n is the number of points and d is the dimension of the point cloud \citep{roadmap}.
